GMSE: an R package for generalised management strategy evaluation

### 2017-11-10 notebook/ms/SI1.Rmd
More edits

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
linestretch: 1
subtitle: Supporting Information 1
header-includes:
- \usepackage{lineno}
- \linenumbers
bibliography: gmse.bib
biblio-style: apalike
---

@@

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework). We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

<a name="Key_data_structures_used">Key data structures used
================================================================================

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.3.1.9 manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

@@

```{r, echo = FALSE}
samp_COST <- sim$COST;
dimnames(samp_COST)[[1]] <- c("Resource", "Landscape", "Res_cost", "U1_cost", "U2_cost");
dimnames(samp_COST)[[2]] <- c("Act", "Type_1", "Type_2", "Type_3", "Util.", "U_land", "U_loc.", "Scare", "Cull", "Castrate", "Feed", "Help_off", "None");
dimnames(samp_COST)[[3]] <- c("Manager_Actions", "User_1_Actions", "User_2_Actions");
print(samp_COST);
```

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are set to `FALSE`. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users. The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.

@@

Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array. The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied. The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies). If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).

@@

Given the complexity of adaptive management and socio-ecological interactions, the above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions.

@@

$$\Delta L_{i} = (\phi)Tends - Kills + r\Delta A_{i}.$$

That is, the change in landscape output equals the increase in output from tending crops, minus the number of crops destroyed, plus the change in resource abundance times the effect that resource abundance has on landscape output.

>>>>

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie, Jeremy J. Cusack, Isabel L. Jones, Erlend B. Nilsen, Roc&#0237;o
  Pozo, O. Sarobidy Rakotonarivo, Bram Van Moorter, and Nils Bunnefeld
date: '`r Sys.Date()`'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
- \usepackage{lineno}
- \linenumbers
linestretch: 1
bibliography: gmse.bib
subtitle: Supporting Information 1
biblio-style: apalike
---

@@

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy-setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users copy). Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework, and we highlight some of these potential avenues for expansion). We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

<a name="Key_data_structures_used">Key data structures used</a>
================================================================================

@@

Columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in predefined GMSE v0.3.1.9 manager and user sub-models (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the *current* manager's utility for resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the most recent estimate of resource abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse` and `gmse_apply`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

@@

```{r, echo = FALSE}
samp_COST <- sim$COST;
dimnames(samp_COST)[[1]] <- c("Resource", "Landscape", "Res_cost", "U1_cost", "U2_cost");
dimnames(samp_COST)[[2]] <- c("Act", "Type_1", "Type_2", "Type_3", "Util.", "U_land", "U_loc.", "Scare", "Cull", "Castrate", "Feed", "Help_off", "None");
dimnames(samp_COST)[[3]] <- c("Manager_Costs", "User_1_Costs", "User_2_Costs");
print(samp_COST);
```

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are set to `FALSE`. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of culling to `r samp_ACTION[3,9,1]` (row 3), and the corresponding `COST` of resource culling is `r samp_COST[1,9,2]` for both users (row 1). The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.

@@

Once a new set of `ga_popsize` individuals is retained through tournament selection, these individuals replace the previous `POPULATION` array. The genetic algorithm terminates if and only if a minimum number of generations has passed (`ga_mingen`) and a convergence criteria (`converge_crit`) is satisfied. The convergence criteria checks the difference between the mean fitness of individuals in the new generation versus the previous generation; if this difference is greater than `converge_crit`, then termination does not occur (this prevents termination from occuring while fitness is still increasing, though it is usually fine to use the default GMSE `converge_crit = 100` and `ga_mingen = 40`, which nearly always terminates the genetic algorithm after 40 generations having identified adaptive manager or user strategies). Due to the way in which fitness is calculated (see below), in practice, `converge_crit` currently applies only to users. If termination conditions are not satisfied, then the `POPULATION` of individuals begins a new generation of crossover, mutation, cost constraint, fitness evaluation, and tournament selection (Figure S1-1).

@@

Given the complexity of adaptive management and socio-ecological interactions, the above conditions for applying the metaheuristic approach are clearly satisfied for manager and user decisions. With this in mind, we now explain the details of manager and user fitness functions; that is, how GMSE assesses whether or not a strategy is a good one.

@@

$$\Delta L_{i} = (\phi)Tends - Kills - r\Delta A_{i}.$$

That is, the change in landscape output equals the increase in output from tending crops, minus the number of crops destroyed, minus the change in resource abundance times the effect that resource abundance has on landscape output (note that if user actions decrease resource abundance, then this last term will be positive, increasing landscape output).
